\documentclass[11pt]{penrose}

\input{header}

\title{Past Paper: Partial Differential Equations}
\subtitle{Final Exam, 2022}
\begin{document}

\maketitle
\warningtext

\begin{problem}{1}
    Use an appropriate method to find the solution of the following problem
    \begin{gather*}
        u_t = k u_{xx} - h u, \quad 0 < x < \ell, \quad t > 0,\\
        u(x, 0) = f(x), \quad 0 \leq x \leq \ell,\\
        u_x(0, t) = u_x(\ell, t) = 0, \quad t \geq 0.
    \end{gather*}
    
    \solution
    Using the separation of variables, we look for a non-trivial solution of the form $u(x,t) = X(x)T(t) \neq 0$. As a result, the PDE separates into two ODEs,
    \begin{equation*}
        X\dot{T} = kX''T - hXT
        \implies
        \frac{\dot{T}}{kT} = \frac{X''}{X} - \frac{h}{k}
        \implies
        \frac{\dot{T}}{kT} + \frac{h}{k} = \frac{X''}{X} = \lambda,
    \end{equation*}
    where $\lambda$ is a constant.

    Moreover, the boundary conditions reduce to
    \begin{equation*}
        u_x(0, t) = u_x(\ell, t) = 0
        \implies
        X'(0) = X'(\ell) = 0.
    \end{equation*}

    Solving the ODE for $T(t)$ leads to,
    \begin{equation*}
        \frac{\dot{T}}{kT} + \frac{h}{k} = \lambda
        \implies
        \frac{\dot{T}}{T} = k\lambda - h
        \implies
        \ln T = -(h - k\lambda)t + \mu
        \implies
        T(t) = A e^{-(h - k\lambda)t},
    \end{equation*}
    where $\mu$ and $A$ are constants.

    Next, we consider $X'' - \lambda X = 0$. This leads to three separate cases.

    \underline{$\lambda > 0$:} The auxiliary equation is then $m^2 - \lambda = 0$. So, $m = \pm \sqrt{\lambda}$ and we get
    \begin{equation*}
        X(x) = c_1 \sinh(\sqrt{\lambda}x) + c_2 \cosh(\sqrt{\lambda}x).
    \end{equation*}
    So, $X'(0) = 0$ implies $c_1 = 0$ and then $X'(\ell) = 0$ requires $c_2 = 0$. This is the trivial solution $X \equiv 0$, so we ignore it.

    \underline{$\lambda = 0$:} The ODE simplifies to $X'' = 0$. So, $X(x) = c_0 + c_1 x$. Now, $X'(0) = 0$ and $X'(\ell) = 0$ both imply $c_1 = 0$. Therefore, $X_0(x) = c_0$, a constant.

    \underline{$\lambda < 0$:} Let $\lambda = - \alpha^2$, then the auxiliary equation is $m^2 + \alpha^2 = 0$. So, $m = \pm i\alpha$ and we get
    \begin{equation*}
        X(x) = c_1 \sin(\alpha x) + c_2 \cos(\alpha x).
    \end{equation*}
    So, $X'(0) = 0$ implies $c_1 = 0$ and then $X'(\ell) = 0$ requires $c_2 \sin(\alpha\ell) = 0$. In order to get a non-trivial solution, we put $\alpha\ell = n\pi$ with $n = 1,2,3,\dots$. Overall, $X_n(x) = c_2 \cos(\frac{n\pi}{\ell} x)$.

    Combining these, we get
    \begin{gather*}
        u_0(x,t) = a_0 e^{-ht},\\
        u_n(x,t) = a_n e^{-(h - k(n\pi/\ell)^2)t}.
    \end{gather*}
    By the principle of superposition, the complete solution is
    \begin{equation*}
        u(x,t) = a_0 e^{-ht} + \sum_{n=1}^{\infty} a_n e^{-(h - k(n\pi/\ell)^2)t}.
    \end{equation*}
\end{problem}

\begin{problem}{2}
    Find the Fourier transform of the following functions
    \begin{enumerate}
        \item[(a)] $f_a(x) = \begin{cases} 1, &\abs{x} < a \\ 0, &\abs{x} \geq a \end{cases}$ where $a$ is a constant.
        \item[(b)] $f(x) = \exp(-a \abs{x})$, where $a$ is a constant.
    \end{enumerate}
    
    \solution
    \begin{enumerate}
        \item[(a)] By definition
        \begin{equation*}
            F[f_a(x)]
            = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^{\infty} f_a(x) e^{-ikx} \,dx
            = \frac{1}{\sqrt{2\pi}} \int_{-a}^{a} e^{-ikx} \,dx
            = \frac{1}{\sqrt{2\pi}} \frac{e^{-ikx}}{-ik} \at[\bigg]{-a}^{a}.
        \end{equation*}
        So,
        \begin{equation*}
            F[f_a(x)]
            = \frac{1}{\sqrt{2\pi}} \frac{e^{ika} - e^{-ika}}{ik}\\
            = \sqrt{\frac{2}{\pi}} \frac{\sin{ka}}{k}.
        \end{equation*}
        \item[(b)] $f(x) = \exp(-a \abs{x})$, where $a$ is a constant.
    \end{enumerate}
\end{problem}

\begin{problem}{3}
    Use Fourier transform to determine the solution for the initial value problem
    \begin{gather*}
        u_{tt} = c^2 u_{xx}, \quad -\infty < x < \infty, \quad t > 0,\\
        u(x,0) = f(x), \quad u_{t}(x,0) = g(x), \quad -\infty < x < \infty.
    \end{gather*}
    
    \solution TBC
\end{problem}

\begin{problem}{4}
    What is an exponential order function? Find the Laplace transform of the following functions
    \begin{enumerate}
        \item[(a)] $f(t) = \cos(\omega t)$.
        \item[(a)] $g(t) = H(t-a)$, where $H(t-a)$ denotes the Heaviside or unit-step function.
    \end{enumerate}
    
    \solution TBC
\end{problem}

\begin{problem}{5}
    Solve the one-dimensional wave equation
    \begin{equation*}
        u_{tt} = c^2 u_{xx}, \quad 0 < x < \infty, \quad t > 0,
    \end{equation*}
    with the boundary and initial conditions
    \begin{gather*}
        u(0,t) = A f(t), \quad t \geq 0,\\
        u(x,t) \to 0 \quad\text{as}\quad x \to 0, \quad t \geq 0,\\
        u(x,t) = u_t(x,t) = 0 \quad\text{at}\quad t = 0.
    \end{gather*}
    
    \solution TBC
\end{problem}

\begin{problem}{6}
    Classify the given equation as hyperbolic, parabolic or elliptic PDE. then reduce it into the canonical form
    \begin{equation*}
        u_{xx} + (2 \cosec y) u_{xy} + (\cosec y)^2 u_{yy} = 0.
    \end{equation*}
    
    \solution
    We identify $A = 1$, $B = 2 \cosec y$, $C = \cosec^2 y$, $D = E = 0$. So,
    \begin{equation*}
        B^2 - 4AC = 4 \cosec^2 y - 4 \cosec^2 y = 0.
    \end{equation*}
    Therefore, this is a parabolic PDE.

    The characteristic equation gives
    \begin{equation*}
        \frac{dy}{dx}
        = \frac{B \pm \sqrt{B^2 - 4AC}}{2A}
        = \cosec y
        \implies
        \sin y \,dy = dx
        \implies
        x + \cos y = c,
    \end{equation*}
    for some constant $c$. Let $\alpha = x + \cos y$ and $\beta = x$. Then,
    \begin{align*}
        C^{*} &= A \beta_x^2 + B \beta_x \beta_y + C \beta_y^2 = 1\\
        D^{*} &= A \alpha_{xx} + B \alpha_{xy} + C \alpha_{yy} + D \alpha_{x} + E \alpha_{y} = -\cosec^2 y \cos y = -\cot(y)\cosec(y)\\
        E^{*} &= A \beta_{xx} + B \beta_{xy} + C \beta_{yy} + D \beta_{x} + E \beta_{y} = 0.
    \end{align*}
    and $A^* = B^* = 0$ because this is a parabolic PDE.

    As a result, the transformed PDE is
    \begin{equation*}
        u_{\beta\beta} - \cot(y)\cosec(y) u_\alpha = 0
    \end{equation*}
\end{problem}

\begin{problem}{7}
    Show that $u(x,y) = e^{-x/b} f(ax-by)$ satisfies the equation $b u_x + a u_y = 0$.
    
    \solution TBC
\end{problem}

\begin{problem}{8}
    Find the general solution of the following first order equations
    \begin{enumerate}
        \item[(a)] $x (y^2 + u) u_x - y (x^2 + u) u_y = (x^2 - y^2) u$.
        \item[(b)] $x^2 u_x + y^2 u_y = (x + y) u$.
    \end{enumerate}
    
    \solution
\end{problem}

\end{document}